---
author: Edward F. Hillenaar, MSc. 
categories:
- panvibrational
date: "2022-06-12"
draft: false
excerpt: A Panvibrational Vision on the Nature and Origin of Consciousness
layout: single
subtitle: A Panvibrational Vision on the Nature and Origin of Consciousness
title: Spiritual
external_link: ""
image:
  caption: Holographic Vibration
  focal_point: Smart
links:
- icon: github
  icon_pack: fab
  name: github-data
  url: https://github.com/EdwardHill15/Data.git/
- icon: youtube
  icon_pack: fab
  name: Video
  url: https://www.youtube.com/watch?v=NHMip21Hkp0&t=26s
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(kableExtra)
require(broom)
library(mgcv)
library(ggplot2)
library(wesanderson)
library(psych)
library(corrplot)
library(dplyr)
library(DT)
```

# Introduction

The theoretical foundation of this article is to be found in [@Lowe1990]. Also in [@Segall2019]  and in [@Schilpp1941], [@Whitehead1929], [@Whitehead1933], [@Whitehead1919]. Another reference from the Bibliography [@Coomaraswamy2004]


# Methods

## Data

```{r, echo=FALSE}
library(readxl)
NVM1Onderzoek <- read_excel("D:/Rmarkdown/NVM1Onderzoek.xlsx")
NVM <- as.data.frame(NVM1Onderzoek)
NVM <- NVM[,c(1,4:5,7:11)]
datatable(NVM, extensions = "Buttons", 
          options = list(dom = 'Bfrtip', 
                         buttons = c('copy', 'csv', 'excel',
                         'pdf', 'print')))
```

I am going to use a `dataset` and a *file*. The name of the dataset is **NVM1Onderzoek.xlsx**. 

I have a [link](https://www.totalegezondheid.nl)

# Statistical Analysis

```{r echo=FALSE, table}
names(NVM) <- c("Code", "Gender", "Age", "NEG", "SOM", "VERL", "PSY", "EX")
NVM_des <- describe(NVM[2:8])
#NVM_sum <- summary(NVM[2:8])
NVM_Table1 <- kable(summary(NVM[2:8]), caption = "Summary NVM Database") %>% 
  kable_styling(bootstrap_options = c("bordered", "striped", "hover", "condensed", "responsive"), full_width = FALSE) %>% 
  #kable_material("striped") %>% 
  #kable_material_dark("striped") %>% 
  #kable_minimal("striped") %>% 
  #kable_classic_2(c("hover", "striped"), font_size = 15, html_font = "sans") %>% 
  #row_spec(2:4, italic = T, bold = T, color = "gold", background = "blue") %>% 
  column_spec(4:8, bold = T, color = "gold", background = "blue") %>% 
  column_spec(1:3, bold = T, background = "pink") %>% 
  add_header_above(c("Characteristics" = 3, "NVM-Scores" = 5))
  
```

```{r, echo=FALSE}
library(corrplot)
NVM_var <- (NVM[,c(4:8)]) 
plot_cor_NVM <- print(cor.plot(NVM_var))
```
## Regression Analysis with PSY as outcome variable and SOM, NEG as Predictors

```{r, echo=FALSE}
fit <- lm(PSY~SOM+NEG, data = NVM)
table_regression_NVM <- kable(tidy(fit), digits = 2, caption = "Linear Regression Model for NVM") %>% 
  kable_styling(bootstrap_options = c("bordered", "striped", "hover", "condensed", "responsive"), full_width = F) %>% 

  #kable_material("striped") %>% 
  #kable_material_dark("striped") %>% 
  #kable_minimal("striped") %>% 
  #kable_classic_2(c("hover", "striped"), font_size = 15, html_font = "sans") %>% 
  #row_spec(2:4, italic = T, bold = T, color = "gold", background = "blue") %>% 
  column_spec(1:5, bold = T, color = "gold", background = "blue") %>% 
  add_header_above(c("Linear Regression Coefficients for SOM and NEG" = 5))
table_regression_NVM
```

# GAM model toegepast op de dataframe NVM met Wood's package "mgcv"


```{r echo=FALSE}
NVM_gam <- gam(PSY~s(NEG,SOM,EX,VERL), data = NVM_var) 
vis.gam(NVM_gam, xlab = "predictors", zlab = "PSY outcome", main = "PSY ~ predictors")
```

# Results


# Plots examples
```{r, echo=FALSE}
NVM$Gender[NVM$Gender == "f"] <- "Vrouw"
NVM$Gender[NVM$Gender == "m"] <- "Man"

theme_ed2 <- theme(axis.line.y = element_line(colour = "black", size = 0.1),
                     axis.line.x = element_line(colour = "black", size = 0.1),
                     panel.grid.minor = element_blank(),
                     panel.background = element_rect(fill="grey95"),
                     panel.border = element_rect(colour = "black", fill = NA),
                     axis.title.x = element_text(size = 20, margin = margin(5,0,0,0)),
                     axis.title.y = element_text(size = 20, margin = margin(0,10,0,0)),
                     axis.text = element_text(size = 16, colour = "black"),
                     axis.text.x = element_text(margin = margin(t=5)),
                     plot.title = element_text(size = 32, hjust = 0),
                     strip.text = element_text(size = 14, face = "bold"),
                     strip.background = element_rect(colour = "black"),
                     panel.spacing = unit(0,"lines")
)

bp <- ggplot(NVM, aes(x=Gender, y=SOM)) +
  geom_boxplot(fill='#A4A4A4', color="darkred")


bp2 <- ggplot(NVM, aes(x=Gender, y=EX, fill=Gender)) +
  geom_point(position = position_jitter(0.2), size = 2, shape = 21) +
  geom_boxplot(alpha=0.8, notch = TRUE)

#bp
#bp + scale_fill_hue(l=40, c=35)
#bp + scale_fill_manual(values=c("#999999", "#E69F00"))
#bp
#bp + scale_fill_manual(breaks = c("2", "1"), 
 #                      values=c("red", "blue"))

#bp + scale_fill_brewer(palette="Blues")

#bp+scale_fill_manual(values=wes_palette(n=3, name="Moonrise2"))

bp + scale_fill_grey() + theme_classic()
bp2 + scale_fill_grey(start=0.8, end=0.5) + theme_classic() + theme_ed2 + ggtitle("Exraversie naar Geslacht") + 
  theme(plot.title = element_text(hjust = 0.5))

```

# SCL dataframe

# Summary and Conclusions

# References


