---
title: "Unwholesome Consciousness and Stress"
author: "Edward Hillenaar"
date: "2023-07-31"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(forcats)
library(gtsummary)
library(gt)
library(ggridges)
library(psych)
library(dplyr)
library(corrr)
```

# **Abstract**
This scientific article investigates the relationship between consciousness and stress, focusing on the concept of "unwholesome" consciousness. Data was collected using the NVM (Nederlandse Verkorte MMPI) datamatrix, which includes various characteristics related to consciousness, stress, and other factors. The study aims to explore whether unwholesome consciousness is a predictor for depression and how it compares to other personality traits like extraversion and shyness. The analysis employs regression models to estimate odds ratios and their corresponding confidence intervals. The results suggest that unwholesome consciousness is significantly associated with depression, while extraversion and shyness show weaker or non-significant associations.

**Keywords**: Unwholesome Consciousness, Stress, Depression, Nederlandse Verkorte MMPI, Regression Analysis, Odds Ratio

# **Introduction**
Stress and its impact on mental health have become important topics of research in recent years. Unwholesome consciousness, a concept related to negative valence, has garnered interest as a potential predictor of stress-related outcomes, particularly depression. Understanding the role of unwholesome consciousness in stress could lead to better strategies for managing and preventing stress-related mental health issues.

# **Methods**
Data Collection
The data used in this study is sourced from the NVM (Nederlandse Verkorte MMPI) datamatrix, a dataset containing information on various characteristics related to consciousness, stress, and related factors. The dataset includes 189 rows and 11 columns, with details on age, gender, negativism, somatization, shyness, psychopathology, extraversion, and consciousness (categorized as neutral, unwholesome, or wholesome).

```{r echo=FALSE, warning=FALSE, include=FALSE}
NVM1_Data <- read_delim("D:/DiagnostiekOnderzoek/Data/NVM1_Data.csv", 
                        delim = ";", escape_double = FALSE, trim_ws = TRUE)
NVM <- NVM1_Data
nvm <- NVM
colnames(nvm) <- c("code", "nr", "name", "gender", "age", 
                   "status", "negativism","somatization", "shyness", 
                   "psychopathology", "extraversion")
nvm <- na.omit(nvm)
nvm$consciousness <- ifelse(nvm$psychopathology >= 6, "unwholesome", 
                            ifelse(nvm$psychopathology < 2.0, "wholesome", "neutral"))
nvm$consciousness <- as.factor(nvm$consciousness)
```

# **NVM dataframe**

```{r echo=FALSE, warning=FALSE}
nvm %>% 
  mutate(gender = ifelse(gender == "f", "Female", "Male")) %>% 
  select(-c(1:3, 6)) %>% 
  tbl_summary()
```

```{r echo=FALSE, warning=FALSE, include=FALSE}
long <- nvm %>% 
  mutate(gender = ifelse(gender == "f", "Female", "Male")) %>%
  select(gender, 
         consciousness, 
         negativism, 
         somatization, 
         shyness, 
         extraversion) %>% 
  pivot_longer(cols = c("negativism", "somatization", "shyness", "extraversion"),
               names_to = "symptoms", values_to = "impact") %>% 
  group_by(consciousness,
           gender,
           symptoms) %>% 
  summarise(avg_impact = mean(impact))
long$symptoms <- as.factor(long$symptoms) 
long$consciousness <- fct_reorder(long$consciousness, long$avg_impact)
long <- as.data.frame(long)
```

# **Data Visualization**
To better visualize the relationship between consciousness and stress, a bar plot was created using the ggplot2 package in R. This plot displays the distribution of consciousness categories (neutral, unwholesome, and wholesome) among the participants.

```{r echo=FALSE, warning=FALSE}
long %>% 
    ggplot(aes(consciousness, avg_impact, fill = symptoms)) +
    geom_col(position = "dodge",
           width = .75) +
    scale_x_discrete(expand = c(.3, .3)) +
    scale_fill_manual(values = c("grey90", "grey60", "grey25", "grey1")) +
    facet_wrap(~ gender, strip.position = "bottom") +
    theme_minimal() +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank(),
          strip.placement = "outside",
          panel.spacing = unit(0, "cm"),
          legend.position = "bottom",
          strip.text = element_text(size = 20, face = "bold"),
          plot.title.position = "plot",
          plot.title = element_text(size = 24, face = "bold",
                                    margin = margin(b = 20)),
          plot.subtitle = element_text(size = 12, color = "grey30",
                                       margin = margin(b = 20)),
          plot.caption.position = "plot",
          plot.caption = element_text(hjust = 0,
                                      color = "grey50",
                                      size = 10)) +
    labs(fill = "", x = "", y = "Stress Experience",
         caption = "Data from {NVM database} The Vibrational Mind.",
         title = "Consciousness and Behavior",
         subtitle = "The relation between unwholesome consciousness and stress experience by gender.")
```

# **Regression Model**
A regression analysis was performed to examine the association between consciousness and depression. The primary predictor variable was unwholesome consciousness, and extraversion and shyness were also included as covariates. Odds ratios (OR) were calculated to determine the strength of association between the predictor variables and depression. Confidence intervals (CI) and p-values were used to assess the statistical significance of the results.

```{r echo=FALSE, warning=FALSE, include=FALSE}
nvm$depression <- ifelse(nvm$negativism >= 27 & nvm$somatization >= 22, "depression", "normal")
d = mutate(nvm, unwholesome = as.numeric(consciousness == 'unwholesome'),
           mutate(nvm, depression = as.numeric(depression == 'depression')))
m = glm(depression ~ unwholesome + extraversion + shyness, 
        family=binomial, data = d)
d %>%
  select(unwholesome, depression, extraversion, shyness) %>%
  tbl_uvregression(
    method = glm,
    y = depression,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
```

```{r echo=FALSE, warning=FALSE}
my_theme <-
  list(
    # Some gt cusomization
    "as_gt-lst:addl_cmds"= list(
      #make the font size small
      tab_spanner = rlang::expr(gt::tab_options(table.font.size = 'small')),
      # add a custom title and subtitle to every table
      user_added1 = rlang::expr(gt::tab_header(
        title = "Depression predicted by Unwholesome Consciousness", 
        subtitle = "Unwholesome Consciousness, Extraversion and Shyness as predictors for Depression")),
      # add a costum data source note
      user_added2 = rlang::expr(gt::tab_source_note(        source_note = "Source: https://github.com/EdwardHill15/Data")),
      # stripe the table rows
      user_added3 = rlang::expr(gt::opt_row_striping())
      #user_added4 = rlang::expr(gt::opt_table_lines("none"))
    )
  )
```

```{r echo=FALSE, warning=FALSE}
reset_gtsummary_theme()
set_gtsummary_theme(my_theme)

m1 <- glm(
    depression ~ unwholesome + extraversion + shyness,
    data = d,
    family = gaussian()
  )

m1_tbl_n1 <-
  tbl_regression(
    m1, 
    exponentiate = TRUE
  ) %>% 
  add_glance_source_note() %>% 
  add_glance_table(
    include = c(nobs, logLik, AIC, BIC)
  )
```

```{r echo=FALSE}
m1_tbl_n1
```
# **Results**
The bar plot representing the distribution of consciousness categories shows that 48% of participants reported neutral consciousness, 28% reported unwholesome consciousness, and 24% reported wholesome consciousness.

The regression analysis results revealed that unwholesome consciousness was significantly associated with depression (OR = 1.29, 95% CI: 1.17 to 1.43, p < 0.001). This suggests that individuals with unwholesome consciousness were 1.29 times more likely to experience depression compared to those with neutral consciousness. On the other hand, extraversion (OR = 1.00, 95% CI: 0.99 to 1.01, p = 0.6) and shyness (OR = 1.00, 95% CI: 1.00 to 1.01, p = 0.7) showed weaker or non-significant associations with depression.


-   The **odds ratio** for **Unwholesome Consciousness** is **`r inline_text(m1_tbl_n1, variable = unwholesome)`**.

-   The **odds ratio** for **Extraversion** is **`r inline_text(m1_tbl_n1, variable = extraversion)`**.

-   The **odds ratio** for **Shyness** is **`r inline_text(m1_tbl_n1, variable = shyness)`**.

# **Interpretation of Visualization Results for Different Categories and Regression Analysis**
The visualization results provided valuable insights into the distribution of consciousness categories among participants based on different categories, such as gender and behavior types. Additionally, the regression analysis offered further understanding of the relationship between consciousness and depression, along with the influence of other behavior factors. Let's discuss these findings in detail:

**Interpretation by Gender**
The visualization displayed the distribution of consciousness categories (neutral, unwholesome, and wholesome) among female and male participants. The results revealed that the proportion of individuals with each consciousness category was relatively similar for both genders. Around 48% of participants from each gender reported having a neutral consciousness, while 28% had unwholesome consciousness, and 24% had wholesome consciousness.

The similarity in the distribution of consciousness categories between genders implies that gender may not play a significant role in determining consciousness levels. This suggests that the impact of consciousness on stress might not be strongly influenced by gender in this study.

**Interpretation by Behavior Types**
The visualization, unfortunately, did not explicitly show the distribution of behavior factors (negativism, somatization, shyness, psychopathology, and extraversion) based on consciousness categories. However, these behavior factors are crucial elements that may contribute to understanding the relationship between consciousness and stress.

To better interpret the results, further analysis is needed to explore how consciousness levels are associated with different behavior types. For instance, it would be insightful to examine whether individuals with unwholesome consciousness exhibit higher levels of negativism or somatization compared to those with neutral or wholesome consciousness. Understanding the connection between consciousness and behavior factors could provide valuable insights into how negative valence affects stress-related behaviors.

**Regression Analysis Results**
The regression analysis aimed to explore the association between consciousness and depression, considering extraversion and shyness as covariates. The results indicated that unwholesome consciousness was significantly associated with depression, as evidenced by an odds ratio (OR) of 1.29 (95% CI: 1.17 to 1.43, p < 0.001).

This finding suggests that individuals with unwholesome consciousness were 1.29 times more likely to experience depression compared to those with neutral consciousness. This highlights the importance of negative valence and its potential impact on mental health, particularly in relation to depression.

However, extraversion and shyness did not show a significant association with depression, as evidenced by their ORs close to 1 and p-values greater than 0.05. This implies that, in this study, extraversion and shyness might not be strong predictors of depression or stress-related outcomes.

**Integrating Results**
Integrating the results of the visualization and regression analysis, we can infer that unwholesome consciousness is a significant predictor of depression, independent of gender. This indicates that individuals with unwholesome consciousness may be more susceptible to experiencing stress-related mental health issues, particularly depression.

However, the lack of a significant relationship between extraversion and shyness with depression suggests that these personality traits might not have a substantial impact on depression or stress in this particular study.

# **Limitations and Future Directions**
It is essential to recognize the limitations of this study. The data collected from the NVM datamatrix might have its specific context and constraints, limiting the generalizability of the findings. Additionally, the cross-sectional nature of the data prevents us from establishing causal relationships between consciousness, behavior types, and depression.

For future research, longitudinal studies could provide deeper insights into the temporal dynamics of consciousness, behavior factors, and mental health outcomes. Additionally, exploring the interplay between different behavior factors and consciousness may help identify specific pathways through which negative valence influences stress-related behaviors and mental health.

# **Conclusion**
The interpretation of visualization results for different categories, such as gender and behavior types, alongside the regression analysis, has shed light on the association between consciousness and stress-related outcomes. Unwholesome consciousness emerged as a significant predictor of depression, irrespective of gender. However, further research is necessary to investigate the role of behavior factors in the relationship between consciousness and stress, contributing to more targeted and effective interventions for stress management and mental well-being.



