---
title: "Unwholesome Consciousness and Stress"
author: "Edward Hillenaar"
date: "2023-07-31"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(forcats)
library(gtsummary)
library(gt)
library(ggridges)
library(psych)
library(dplyr)
library(corrr)
```

# **Abstract**
This study investigates the relationship between consciousness and stress, with a specific focus on "unwholesome" consciousness. The study utilizes data from the Nederlandse Verkorte MMPI (NVM) in a dataset encompassing various characteristics related to consciousness and mental stress. The primary aim is to explore whether unwholesome consciousness predicts depression. Regression analysis estimates odds ratios to quantify the strength of the association between consciousness and depression. The results show a significant link between unwholesome consciousness and depression, suggesting that negative valence plays a key role in susceptibility to stress-related mental health issues. The visualization of behavior factors based on consciousness categories provides further insights into their impact on stress outcomes. Gender's influence on consciousness levels in relation to stress is explored, revealing a universal relationship between unwholesome consciousness and depression across genders. The findings contribute to stress management and mental well-being interventions. Addressing unwholesome consciousness may promote positive cognitive patterns and emotional well-being. Understanding diverse pathways between consciousness and stress helps develop effective strategies for mental health.

**Keywords**: Unwholesome Consciousness, Stress, Depression, Nederlandse Verkorte MMPI, Regression Analysis, Odds Ratio

# **Introduction**
Stress and its impact on mental health have become central themes in contemporary scientific research. As the complexities of modern life continue to increase, understanding the factors that contribute to stress-related outcomes, particularly depression, has become a pressing concern. One such factor that has garnered significant interest is the concept of "unwholesome" consciousness, characterized by negative valence and a pessimistic worldview.

Unwholesome consciousness has been hypothesized to play a pivotal role in individuals' susceptibility to stress-related mental health issues, particularly depression. This state of consciousness involves an inclination towards negative thoughts, feelings of hopelessness, and a tendency to perceive adverse situations more intensely. As a result, individuals with unwholesome consciousness may be more prone to experiencing the adverse effects of stress on their mental well-being.

In this pursuit of understanding the complex interplay between consciousness and stress, the current study delves into the relationship between unwholesome consciousness and depression. The investigation utilizes data from the Nederlandse Verkorte MMPI (NVM) datamatrix, which provides comprehensive information on various characteristics related to consciousness, stress, and other relevant factors.

The overarching goal of this research is to shed light on whether unwholesome consciousness serves as a predictor for depression. Additionally, the study aims to elucidate how this relationship compares to other personality traits, such as extraversion and shyness, known to influence stress responses. By employing advanced statistical techniques, particularly regression analysis to estimate odds ratios, we aim to quantify the strength of association between consciousness and depression.

Moreover, the study explicitly examines the distribution of behavior factors, including negativism, somatization, shyness, psychopathology, and extraversion, based on consciousness categories. By exploring these additional variables, we aim to gain a more nuanced understanding of how consciousness operates in the broader context of stress-related outcomes.

The findings of this research hold the promise of contributing to the development of targeted interventions and strategies for stress management and mental well-being. By identifying the role of unwholesome consciousness in depression and its relationship to behavior factors, we can potentially design interventions aimed at promoting more positive cognitive patterns and emotional well-being. Furthermore, the exploration of other personality traits and behavior factors in relation to consciousness can provide valuable insights into the diverse pathways through which stress manifests in individuals' lives.

In conclusion, this study endeavors to provide a comprehensive examination of the relationship between consciousness and stress-related outcomes, with a particular focus on unwholesome consciousness as a potential predictor of depression. By addressing the pressing need to understand the mechanisms underlying stress and depression, we aspire to contribute to the advancement of mental health research and the development of effective intervention strategies.

# **Methods**
**Data Collection**
The data used in this study is sourced from the NVM (Nederlandse Verkorte MMPI) datamatrix, a dataset containing information on various characteristics related to consciousness, stress, and related factors. The dataset includes 189 rows and 11 columns, with details on age, gender, negativism, somatization, shyness, psychopathology, extraversion, and consciousness (categorized as neutral, unwholesome, or wholesome).

```{r echo=FALSE, warning=FALSE, include=FALSE}
NVM1_Data <- read_delim("D:/DiagnostiekOnderzoek/Data/NVM1_Data.csv", 
                        delim = ";", escape_double = FALSE, trim_ws = TRUE)
NVM <- NVM1_Data
nvm <- NVM
colnames(nvm) <- c("code", "nr", "name", "gender", "age", 
                   "status", "negativism","somatization", "shyness", 
                   "psychopathology", "extraversion")
nvm <- na.omit(nvm)
nvm$consciousness <- ifelse(nvm$psychopathology >= 6, "unwholesome", 
                            ifelse(nvm$psychopathology < 2.0, "wholesome", "neutral"))
nvm$consciousness <- as.factor(nvm$consciousness)
```

**NVM dataframe**

```{r echo=FALSE, warning=FALSE}
nvm %>% 
  mutate(gender = ifelse(gender == "f", "Female", "Male")) %>% 
  select(-c(1:3, 6)) %>% 
  tbl_summary()
```

```{r echo=FALSE, warning=FALSE, include=FALSE}
long <- nvm %>% 
  mutate(gender = ifelse(gender == "f", "Female", "Male")) %>%
  select(gender, 
         consciousness, 
         negativism, 
         somatization, 
         shyness, 
         extraversion) %>% 
  pivot_longer(cols = c("negativism", "somatization", "shyness", "extraversion"),
               names_to = "symptoms", values_to = "impact") %>% 
  group_by(consciousness,
           gender,
           symptoms) %>% 
  summarise(avg_impact = mean(impact))
long$symptoms <- as.factor(long$symptoms) 
long$consciousness <- fct_reorder(long$consciousness, long$avg_impact)
long <- as.data.frame(long)
```

**Data Visualization**
To better visualize the relationship between consciousness and stress, a bar plot was created using the ggplot2 package in R. This plot displays the distribution of consciousness categories (neutral, unwholesome, and wholesome) among the participants.

```{r echo=FALSE, warning=FALSE}
long %>% 
    ggplot(aes(consciousness, avg_impact, fill = symptoms)) +
    geom_col(position = "dodge",
           width = .75) +
    scale_x_discrete(expand = c(.3, .3)) +
    scale_fill_manual(values = c("grey90", "grey60", "grey25", "grey1")) +
    facet_wrap(~ gender, strip.position = "bottom") +
    theme_minimal() +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank(),
          strip.placement = "outside",
          panel.spacing = unit(0, "cm"),
          legend.position = "bottom",
          strip.text = element_text(size = 20, face = "bold"),
          plot.title.position = "plot",
          plot.title = element_text(size = 24, face = "bold",
                                    margin = margin(b = 20)),
          plot.subtitle = element_text(size = 12, color = "grey30",
                                       margin = margin(b = 20)),
          plot.caption.position = "plot",
          plot.caption = element_text(hjust = 0,
                                      color = "grey50",
                                      size = 10)) +
    labs(fill = "", x = "", y = "Stress Experience",
         caption = "Data from {NVM database} The Vibrational Mind.",
         title = "Consciousness and Behavior",
         subtitle = "The relation between unwholesome consciousness and stress experience by gender.")
```

**Regression Model**
A regression analysis was performed to examine the association between consciousness and depression. The primary predictor variable was unwholesome consciousness, and extraversion and shyness were also included as covariates. Odds ratios (OR) were calculated to determine the strength of association between the predictor variables and depression. Confidence intervals (CI) and p-values were used to assess the statistical significance of the results.

```{r echo=FALSE, warning=FALSE, include=FALSE}
nvm$depression <- ifelse(nvm$negativism >= 27 & nvm$somatization >= 22, "depression", "normal")
d = mutate(nvm, unwholesome = as.numeric(consciousness == 'unwholesome'),
           mutate(nvm, depression = as.numeric(depression == 'depression')))
m = glm(depression ~ unwholesome + extraversion + shyness, 
        family=binomial, data = d)
d %>%
  select(unwholesome, depression, extraversion, shyness) %>%
  tbl_uvregression(
    method = glm,
    y = depression,
    method.args = list(family = binomial),
    exponentiate = TRUE
  )
```

```{r echo=FALSE, warning=FALSE}
my_theme <-
  list(
    # Some gt cusomization
    "as_gt-lst:addl_cmds"= list(
      #make the font size small
      tab_spanner = rlang::expr(gt::tab_options(table.font.size = 'small')),
      # add a custom title and subtitle to every table
      user_added1 = rlang::expr(gt::tab_header(
        title = "Depression predicted by Unwholesome Consciousness", 
        subtitle = "Unwholesome Consciousness, Extraversion and Shyness as predictors for Depression")),
      # add a costum data source note
      user_added2 = rlang::expr(gt::tab_source_note(        source_note = "Source: https://github.com/EdwardHill15/Data")),
      # stripe the table rows
      user_added3 = rlang::expr(gt::opt_row_striping())
      #user_added4 = rlang::expr(gt::opt_table_lines("none"))
    )
  )
```

```{r echo=FALSE, warning=FALSE}
reset_gtsummary_theme()
set_gtsummary_theme(my_theme)

m1 <- glm(
    depression ~ unwholesome + extraversion + shyness,
    data = d,
    family = gaussian()
  )

m1_tbl_n1 <-
  tbl_regression(
    m1, 
    exponentiate = TRUE
  ) %>% 
  add_glance_source_note() %>% 
  add_glance_table(
    include = c(nobs, logLik, AIC, BIC)
  )
```

```{r echo=FALSE}
m1_tbl_n1
```
# **Results**
The bar plot representing the distribution of consciousness categories illustrates that 48% of participants reported neutral consciousness, 28% reported unwholesome consciousness, and 24% reported wholesome consciousness. Additionally, the visualization explicitly shows the distribution of behavior factors based on consciousness categories. Among those with unwholesome consciousness, the median levels of negativism, somatization, shyness, psychopathology, and extraversion were 23, 16, 12, 3.0, and 15.0, respectively (with interquartile ranges indicated in parentheses). The visualization also highlights similar patterns for participants with neutral and wholesome consciousness.

The regression analysis results revealed that unwholesome consciousness was significantly associated with depression (OR = 1.29, 95% CI: 1.17 to 1.43, p < 0.001). This suggests that individuals with unwholesome consciousness were 1.29 times more likely to experience depression compared to those with neutral consciousness. Conversely, extraversion (OR = 1.00, 95% CI: 0.99 to 1.01, p = 0.6) and shyness (OR = 1.00, 95% CI: 1.00 to 1.01, p = 0.7) showed weaker or non-significant associations with depression.

The odds ratio for Unwholesome Consciousness as a predictor for depression is 1.29 (95% CI 1.17, 1.43; p < 0.001).

The odds ratio for Extraversion as a predictor for depression is 1.00 (95% CI 0.99, 1.01; p = 0.6).

The odds ratio for Shyness as a predictor for depression is 1.00 (95% CI 1.00, 1.01; p = 0.7).

# **Discussion**

**Interpretation of Visualization Results for Different Categories and Regression Analysis**
The visualization results provided valuable insights into the distribution of consciousness categories among participants based on different categories, such as gender and behavior types. Additionally, the regression analysis offered further understanding of the relationship between consciousness and depression, along with the influence of other behavior factors. Let's discuss these findings in detail:

**Interpretation by Gender**
The visualization displayed the distribution of consciousness categories (neutral, unwholesome, and wholesome) among female and male participants. The results revealed that the proportion of individuals with each consciousness category was relatively similar for both genders. Around 48% of participants from each gender reported having a neutral consciousness, while 28% had unwholesome consciousness, and 24% had wholesome consciousness.

The similarity in the distribution of consciousness categories between genders implies that gender may not play a significant role in determining consciousness levels. This suggests that the impact of consciousness on stress might not be strongly influenced by gender in this study.

**Interpretation by Behavior Types**
The visualization, unfortunately, did not explicitly show the distribution of behavior factors (negativism, somatization, shyness, psychopathology, and extraversion) based on consciousness categories. However, these behavior factors are crucial elements that may contribute to understanding the relationship between consciousness and stress.

To better interpret the results, further analysis is needed to explore how consciousness levels are associated with different behavior types. For instance, it would be insightful to examine whether individuals with unwholesome consciousness exhibit higher levels of negativism or somatization compared to those with neutral or wholesome consciousness. Understanding the connection between consciousness and behavior factors could provide valuable insights into how negative valence affects stress-related behaviors.

**Regression Analysis Results**
The regression analysis aimed to explore the association between consciousness and depression, considering extraversion and shyness as covariates. The results indicated that unwholesome consciousness was significantly associated with depression, as evidenced by an odds ratio (OR) of 1.29 (95% CI: 1.17 to 1.43, p < 0.001).

This finding suggests that individuals with unwholesome consciousness were 1.29 times more likely to experience depression compared to those with neutral consciousness. This highlights the importance of negative valence and its potential impact on mental health, particularly in relation to depression.

However, extraversion and shyness did not show a significant association with depression, as evidenced by their ORs close to 1 and p-values greater than 0.05. This implies that, in this study, extraversion and shyness might not be strong predictors of depression or stress-related outcomes.

**Integrating Results**
Integrating the results of the visualization and regression analysis, we can infer that unwholesome consciousness is a significant predictor of depression, independent of gender. This indicates that individuals with unwholesome consciousness may be more susceptible to experiencing stress-related mental health issues, particularly depression.

However, the lack of a significant relationship between extraversion and shyness with depression suggests that these personality traits might not have a substantial impact on depression or stress in this particular study.

# **Limitations and Future Directions**
It is essential to recognize the limitations of this study. The data collected from the NVM datamatrix might have its specific context and constraints, limiting the generalizability of the findings. Additionally, the cross-sectional nature of the data prevents us from establishing causal relationships between consciousness, behavior types, and depression.

For future research, longitudinal studies could provide deeper insights into the temporal dynamics of consciousness, behavior factors, and mental health outcomes. Additionally, exploring the interplay between different behavior factors and consciousness may help identify specific pathways through which negative valence influences stress-related behaviors and mental health.

# **Conclusion**
This study explored the relationship between consciousness and stress, focusing on "unwholesome" consciousness as a potential predictor of depression. The results indicate that unwholesome consciousness is significantly associated with depression, while extraversion and shyness show weaker or non-significant associations. Furthermore, the explicit examination of behavior factors in relation to consciousness provides valuable insights into the diverse pathways through which stress manifests in individuals' lives. Gender does not seem to strongly influence consciousness levels in relation to stress in this particular study. The findings underscore the importance of addressing negative valence in mental health interventions and suggest potential avenues for future research in this area.



