---
author: Edward F. Hillenaar, MSc. 
categories:
- temperature
date: "2022-07-22"
draft: false
excerpt: Temperature rises in the Netherlands between 1901 and 2022
layout: single
subtitle: Temperature rises in the Netherlands between 1901 and 2022
title: Temperature Rise in The Netherlands
external_link: ""
image:
  caption: Temprature
  focal_point: Smart
links:
- icon: github
  icon_pack: fab
  name: github-data
  url: https://github.com/EdwardHill15/climate_viz
url_code: ""
url_pdf: ""
url_slides: ""
url_video: ""
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(glue)
library(lubridate)
```

# Introduction

There's a slight temperature rise measured in the Netherlands between 1901 and 2022. Though the maximum temperatures near 37 degrees of Celsius were also measured respectively 15, 30, 50 and 60 years ago. 

Source:1.  Source: KNMI weather station De Bilt in The Netherlands (NLM00006260)
2. https://github.com/EdwardHill15/climate_viz

# Data

## Scraping weather data from the internet with R and the tidyverse (CC231)

* https://riffomonas.org/code_club/2022-07-18-local-weather

## Finding a Weather Station

* https://ncei.noaa.gov/cdo-web/

## Weather Station KNMI De Bilt The Netherlands

* https://www.knmi.nl/home

```{r include=FALSE}
inventory_url <- "https://www.ncei.noaa.gov/pub/data/ghcn/daily/ghcnd-inventory.txt"

inventory <- read_table(inventory_url, 
           col_names = c("station", "lat", "lon", "variable", "start", "end"))


my_lat <- 52.13 * 2 * pi / 360
my_lon <- 5.29 * 2 * pi / 360


inventory <- inventory %>% 
  mutate(lat_r = lat*2*pi/360,
         lon_r = lon*2*pi/360,
         d = 3963 * acos((sin(lat_r) * sin(my_lat)) + 
                           cos(lat_r) * cos(my_lat) *
                           cos(my_lon - lon_r)))

```

<!-- # mutate miles into km with * 1.609344 -->
```{r include=FALSE}
my_station <- inventory %>% 
  mutate(lat_r = lat*2*pi/360,
         lon_r = lon*2*pi/360,
         d = 1.609344 * 3963 * acos((sin(lat_r) * sin(my_lat)) + 
                           cos(lat_r) * cos(my_lat) *
                           cos(my_lon - lon_r))) %>% 
  filter(start < 1960 & end > 2020) %>% 
  top_n(n = -1, d) %>% 
  distinct(station) %>% 
  pull(station)
```

## Get and tidy local weather station data (from: "NLE00100501")

```{r include=FALSE}
station_daily <- glue("https://www.ncei.noaa.gov/pub/data/ghcn/daily/by_station/{my_station}.csv.gz")

df <- read_csv(station_daily,
         col_names = c("station", "date", "variable", "value", "a", "b", "c", "d")) %>% 
  select(date, variable, value) %>% 
  pivot_wider(names_from = "variable", values_from = "value",
              values_fill = 0) %>% 
  select(date, TMAX, TMIN, PRCP)

```

# Statistical Analysis and plots

## Figure 1: Maximum Temperature in The Netherlands between 1901 and 2022

```{r echo=FALSE}
df <- df %>% 
  mutate(date = ymd(date))
  
NE_max <- aggregate(TMAX~year(date), df, max)
NE_max$TMAX <- NE_max$TMAX / 10
colnames(NE_max) <- c("year", "temp")
NE_max %>% 
  ggplot(aes(x=year, y=temp, color = temp)) +
  geom_point(size = 9, alpha = .7) +
  geom_smooth(method = "lm", formula = y ~x, se = FALSE, color = "gray") +
  scale_color_viridis_c(option = ("magma")) +
  # scale_color_gradient2(low = "blue", mid = "white", high = "red",
  #                       midpoint = 13.5) + 
#coord_cartesian(xlim=c(10,16)) +
#coord_cartesian(expand=TRUE) +
  scale_y_continuous(breaks = seq(27.5, 37.5, 2.5),
                   limits = c(26, 37.5)) +
  scale_x_continuous(breaks = seq(1901, 2022, 22),
                     limits = c(1901, 2023)) +
  labs(y = "Temperature in (\u00B0 C)", 
     x = NULL,
     title = "Maximum Temperature The Netherlands",
     subtitle = "Temperature is colored by the size of the maximum temperature value",
     caption = "Source: KNMI weather station De Bilt in The Netherlands (NLM00006260)") +
  theme(
    plot.background = element_rect(fill = "black", color = "black"),
    panel.background = element_rect(fill = "black"),
    plot.title = element_text(color = "white", face = "bold", size = 25),
    plot.subtitle = element_text(color = "gray", size =15),
    plot.title.position = "plot",
    plot.caption = element_text(color = "white", size = 12),
    axis.text = element_text(color = "white", size = 15),
    axis.title = element_text(color = "white", size = 15),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(color = "gray", size = 0.25),
    panel.grid.minor.y = element_blank(),
    axis.ticks = element_blank(),
    legend.background = element_rect(fill = "black"),
    legend.text = element_text(color = "white", size = 6),
    legend.title = element_text(color = "white"),
    legend.position = "bottom")

```

## Figure 2: Mean Temperature in The Netherlands between 1901 and 2022

```{r echo=FALSE}
NE_mean <- aggregate(TMAX~year(date), df, mean)
NE_mean$TMAX <- NE_mean$TMAX / 10
colnames(NE_mean) <- c("year", "temp")
NE_mean$temp <- round(NE_mean$temp, 1)

NE_mean %>% 
  ggplot(aes(x=year, y=temp, color = temp)) +
  geom_point(size = 9, alpha = .7) +
  geom_smooth(method = "lm", formula = y ~x, se = FALSE, color = "gray") +
  scale_color_viridis_c(option = ("viridis")) +
  # scale_color_gradient2(low = "blue", mid = "white", high = "red",
  #                       midpoint = 13.5) + 
  #coord_cartesian(xlim=c(10,16)) +
  #coord_cartesian(expand=TRUE) +
  scale_y_continuous(breaks = seq(11.5, 16, 1.5),
                     limits = c(11.5, 16)) +
  scale_x_continuous(breaks = seq(1901, 2022, 22),
                     limits = c(1901, 2023)) +
  labs(y = "Temperature in (\u00B0 C)", 
       x = NULL,
       title = "Mean Temperature in The Netherlands",
     subtitle = "Temperature is colored by the size of the maximum temperature value",
     caption = "Source: KNMI weather station De Bilt in The Netherlands (NLM00006260)") +
  theme(
    plot.background = element_rect(fill = "black", color = "black"),
    panel.background = element_rect(fill = "black"),
    plot.title = element_text(color = "white", face = "bold", size = 25),
    plot.subtitle = element_text(color = "gray", size =15),
    plot.title.position = "plot",
    plot.caption = element_text(color = "white", size = 12),
    axis.text = element_text(color = "white", size = 15),
    axis.title = element_text(color = "white", size = 15),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(color = "gray", size = 0.25),
    panel.grid.minor.y = element_blank(),
    axis.ticks = element_blank(),
    legend.background = element_rect(fill = "black"),
    legend.text = element_text(color = "white", size = 6),
    legend.title = element_text(color = "white"),
    legend.position = "bottom")
```


## Figure 3: Precipitation in The Netherlands between 1901 and 2022

```{r echo=FALSE}
NE_prcp <- aggregate(PRCP~year(date), df, max)
NE_prcp$PRCP <- NE_prcp$PRCP / 10
colnames(NE_prcp) <- c("year", "prcp")
NE_prcp$prcp <- round(NE_prcp$prcp, 1)
NE_prcp <- NE_prcp[-c(1:5),]
NE_prcp %>% 
  ggplot(aes(x=year, y=prcp, color = prcp)) +
  geom_point(size = 9, alpha = .7) +
  geom_smooth(method = "lm", formula = y ~x, se = FALSE, color = "gray") +
  scale_color_viridis_c(option = ("mako")) +
  # scale_color_gradient2(low = "blue", mid = "white", high = "red",
  #                       midpoint = 13.5) + 
  #coord_cartesian(xlim=c(10,16)) +
  #coord_cartesian(expand=TRUE) +
  scale_y_continuous(breaks = seq(15, 65, 10),
                     limits = c(15, 65)) +
  scale_x_continuous(breaks = seq(1901, 2022, 22),
                     limits = c(1901, 2023)) +
  labs(y = "Precipitation in mm.", 
       x = NULL,
       title = "Precipitation in The Netherlands",
     subtitle = "Temperature is colored by the size of the maximum temperature value",
     caption = "Source: KNMI weather station De Bilt in The Netherlands (NLM00006260)") +
  theme(
    plot.background = element_rect(fill = "black", color = "black"),
    panel.background = element_rect(fill = "black"),
    plot.title = element_text(color = "white", face = "bold", size = 25),
    plot.subtitle = element_text(color = "gray", size =15),
    plot.title.position = "plot",
    plot.caption = element_text(color = "white", size = 12),
    axis.text = element_text(color = "white", size = 15),
    axis.title = element_text(color = "white", size = 15),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(color = "gray", size = 0.25),
    panel.grid.minor.y = element_blank(),
    axis.ticks = element_blank(),
    legend.background = element_rect(fill = "black"),
    legend.text = element_text(color = "white", size = 6),
    legend.title = element_text(color = "white"),
    legend.position = "bottom")
```

# Conclusion

My conclusion based on these data is that there is light positive trend of temperature rise from 1901 to 2022 but this doesn't suppport the idea of a climate-crisis based on measured temperature differences. 